Large and Infinitary Quotient Inductive-Inductive Types

### 2020-01-10 draft/paper.tex
paper

\newcommand{\bCon}{\boldsymbol{\Con}}
\newcommand{\bTy}{\boldsymbol{\Ty}}
\newcommand{\bSub}{\boldsymbol{\Sub}}
\newcommand{\bTm}{\boldsymbol{\Tm}}
\newcommand{\bGamma}{\boldsymbol{\Gamma}}
\newcommand{\bDelta}{\boldsymbol{\Delta}}
\newcommand{\bsigma}{\boldsymbol{\sigma}}
\newcommand{\bdelta}{\boldsymbol{\delta}}
\newcommand{\bepsilon}{\boldsymbol{\epsilon}}
\newcommand{\bt}{\boldsymbol{t}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bA}{\boldsymbol{A}}
\newcommand{\ba}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\bB}{\boldsymbol{B}}
\newcommand{\bid}{\boldsymbol{\id}}
\newcommand{\bemptycon}{\boldsymbol{\emptycon}}
\newcommand{\bSet}{\boldsymbol{\Set}}
\newcommand{\bU}{\boldsymbol{\U}}
\newcommand{\bEl}{\boldsymbol{\El}}
\newcommand{\bPii}{\boldsymbol{\Pi}}
\newcommand{\bPie}{\boldsymbol{\Pie}}
\newcommand{\bPiinf}{\boldsymbol{\Piinf}}
\newcommand{\bappi}{\boldsymbol{\mathsf{app}}}
\newcommand{\blami}{\boldsymbol{\mathsf{lam}}}
\newcommand{\bId}{\boldsymbol{\Id}}
\newcommand{\bM}{\boldsymbol{\mathsf{M}}}
\newcommand{\bT}{\boldsymbol{\mathsf{T}}}
\newcommand{\bS}{\boldsymbol{\mathsf{S}}}

@@

@@

\begin{definition}[The $\Set$ model of $\ToS$] For each $j$, we have $\bS : \ToS_{j+1, j}$, which can be given by restricting the $\bM$ model
of Section \ref{sec:semantics} so that we only have the first $\Con$ components
in the interpretations for contexts, substitution, types, term, and we only have
actions on contexts in the interpretations of term and substitution
formers. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bS} && :\equiv \Set_{j+2} \\
  & \Sub_{\bS}\,\Gamma\,\Delta && :\equiv \Gamma \ra \Delta\\
  & \Ty_{\bS}\,\Gamma && :\equiv \Gamma \ra \Set_{j+2}\\
  & \Tm_{\bS}\,\Gamma\,A && :\equiv (\gamma : \Gamma) \ra A\,\gamma
\end{alignat*}
Now, for $\Gamma : \Con_{\syn}$, the type of $\Gamma$-algebras is given by
$\llb\Gamma\rrb_{\bS}$. E.g.\, $\llb\NatSig\rrb_{\bS}$ yields a left-nested
$\Sigma$-type of pointed sets with an endofunction. Also, $\llb\Gamma\rrb_{\bM}$
extends $\llb\Gamma\rrb_{\bS}$ to an flCwF of $\Gamma$-algebras.
\end{definition}

\begin{definition}[Term algebra construction]
For each syntactic signature $\ulOmega : \SynSig_j$, we define a displayed $\ToS$
model over $\syn_j$, named $\bT$. The underlying sets are as follows:
\begin{alignat*}{3}
  & \Con_{\bT}\,\ulGamma &&:\equiv \Sub\,\ulOmega\,\ulGamma \ra \llb\ulGamma\rrb_{\bS}\\
  & \Sub_{\bT}\,\Gamma\,\Delta\,\ulsigma &&:\equiv (\ulnu : \Sub\,\ulOmega\,\ulGamma)
  \ra \Delta\,(\ulsigma\circ\nu) \simeq \llb\ulsigma\rrb_{\bS}\,(\Gamma\,\ulnu)\\
  & \Ty_{\bT}\,\Gamma\,\ulA &&:\equiv (\ulnu : \Sub\,\ulOmega\,\ulGamma) \ra
    \Tm\,\ulOmega\,(\ulA[\ulnu]) \ra \llb\ulA\rrb_{\bS}\,(\Gamma\,\ulnu)\\
  & \Tm_{\bT}\,\Gamma\,A\,\ult &&:\equiv (\ulnu : \Sub\,\ulOmega\,\ulGamma)\ra
    \llb\ulA\rrb_{\bS}\,\ulnu\,(\ult[\ulnu]) \simeq_{\id} \llb\ult\rrb_{\bS} (\Gamma\,\ulnu)
\end{alignat*}

Above, the $\simeq$ in the definition of $\Sub_{\bT}$ is a context isomorphism
in $\llb\ulDelta\rrb_{\bM}$, which is the flCwF of $\ulDelta$-algebras. The
$\simeq_{\id}$ in $\Ty_{\bT}$ is a vertical context isomorphism in the displayed
flCwF given by $\llb\ulA\rrb_{\bM}$.

So far, the underlying sets in $\bT$ are similar to what was given in
\cite{kaposi2019constructing} in the construction of term algebras, but there is
an important difference: in ibid.\, strict equalities are used instead of
isomorphisms. In our case, isomorphisms are necessary once again because of
infinitary functions types and our identity type; we shall see this shortly. The
universe is interpreted as follows:
\begin{alignat*}{3}
  & \U_{\bT} : (\ulnu : \Sub\,\ulOmega\,\ulGamma)(\ult : \Tm\,\ulOmega\,\U)\ra
              \Set_{j+1}\\
  & \U_{\bT}\,\ulnu\,\ult :\equiv \Tm\,\ulOmega\,(\El\,\ult)\\
  & \El_{\bT}\,a : (\ulnu : \Sub\,\ulOmega\,\ulGamma)
             (\ult : \Tm\,\ulOmega\,(\El\,(\ula[\ulnu])))\ra
              \llb\ula\rrb_{\bS}\,(\Gamma\,\ulnu)\\
  & \El_{\bT}\,a\,\ulnu\,\ult :\equiv (a\,\ulnu)\,\ult
\end{alignat*}

Hence, a syntactic $\ult : \Tm\,\ulOmega\,\U$ is interpreted as a set of terms
with type $\El\,\ult$. In the interpretation of $\El$, note that
\[
  a\,\ulnu : \U_{\bT}\,\ulnu\,(\ula[\ulnu]) \simeq_{\id}
                 \llb\ula\rrb_{\bS}\,(\Gamma\,\ulnu)
\]
hence
\[
a\,\ulnu : \Tm\,\ulOmega\,(\El\,(\ula[\ulnu])) \simeq_{\id}
           \llb\ula\rrb_{\bS}\,(\Gamma\,\ulnu)
\]
The $\simeq_{\id}$ above is just an isomorphism of sets, since it lives in
$\llb\U\rrb_{\bM}$ which was given as the flCwF of sets in Section
\ref{sec:universe}. This above isomorphism is a good summary of the
construction: the interpretation of a $\ula : \Tm\,\ulOmega\,\U$ in the term
algebra is isomorphic to a set of terms.

Inductive functions are interpreted by transport along such isomorphism:
\[
  \Pii_{\bT}\,a\,B\,\ulnu\,\ult :\equiv \lambda\,\alpha.\,
         B\,(\ulnu,\,(a\,\ulnu)^{-1}\,\alpha)\,(\ult\,\appitt\,((a\,\ulnu)^{-1}\,\alpha))
\]

For the infinitary function space, we need the following, where $\simeq_{\id}$ is
again set isomorphism.
\begin{alignat*}{3}
  & \Piinf_{\bT}\,A\,b\,\ulnu : \Tm\,\ulOmega\,(\El\,(\Piinf\,A\,(\lambda\,\alpha.\,(b\,\alpha)[\ulnu])))\\
  & \hspace{4.2em}\simeq_{\id} ((\alpha : A) \ra \llb\alpha\rrb_{\bS}\,(\Gamma\,\ulnu))
\end{alignat*}
This can be given using the natural isomorphism consisting of $\appinf$ and
$\laminf$. However, the sides are not strictly equal. For the identity type, we
build the following isomorphism using equality reflection.
\begin{alignat*}{3}
  & \Id_{\bT}\,a\,t\,u\,\ulnu : \Tm\,\ulOmega\,(\El\,(\Id\,(\ula[\ulnu])\,(\ult[\ulnu])\,(\ulu[\ulnu]))\\
  & \hspace{4.2em}\simeq_{\id} (\llb\ult\rrb_{\bS}\,(\Gamma\,\ulnu) =   \llb\ulu\rrb_{\bS}\,(\Gamma\,\ulnu))
\end{alignat*}

We omit the rest of the definition of $\bT$. The interpretations of equations in
the CwF and fl-structure parts become fairly technical, and we also need to
utilize iso-cleaving to interpret type substitution and substitution
laws. However, the basic shape of the model remains similar to
\cite{kaposi2019constructing}.

@@

%% write about limitations!
p

>>>>

\newcommand{\bCon}{\boldsymbol{\Con}}
\newcommand{\bTy}{\boldsymbol{\Ty}}
\newcommand{\bSub}{\boldsymbol{\Sub}}
\newcommand{\bTm}{\boldsymbol{\Tm}}
\newcommand{\bGamma}{\boldsymbol{\Gamma}}
\newcommand{\bDelta}{\boldsymbol{\Delta}}
\newcommand{\bsigma}{\boldsymbol{\sigma}}
\newcommand{\bdelta}{\boldsymbol{\delta}}
\newcommand{\bepsilon}{\boldsymbol{\epsilon}}
\newcommand{\bt}{\boldsymbol{t}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bA}{\boldsymbol{A}}
\newcommand{\ba}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\bB}{\boldsymbol{B}}
\newcommand{\bid}{\boldsymbol{\id}}
\newcommand{\bemptycon}{\boldsymbol{\emptycon}}
\newcommand{\bSet}{\boldsymbol{\Set}}
\newcommand{\bU}{\boldsymbol{\U}}
\newcommand{\bEl}{\boldsymbol{\El}}
\newcommand{\bPii}{\boldsymbol{\Pi}}
\newcommand{\bPie}{\boldsymbol{\Pie}}
\newcommand{\bPiinf}{\boldsymbol{\Piinf}}
\newcommand{\bappi}{\boldsymbol{\mathsf{app}}}
\newcommand{\blami}{\boldsymbol{\mathsf{lam}}}
\newcommand{\bId}{\boldsymbol{\Id}}
\newcommand{\bM}{\boldsymbol{\mathsf{M}}}
\newcommand{\bT}{\boldsymbol{\mathsf{T}}}
\newcommand{\bS}{\boldsymbol{\mathsf{S}}}
\newcommand{\bP}{\boldsymbol{\mathsf{P}}}
\newcommand{\bD}{\boldsymbol{\mathsf{D}}}

@@

\subsection{Useful Model Fragments of $\bM$}

First, we need to describe three model fragments of $\bM$, which can be
respectively used to compute notions of algebras, displayed algebras and
sections for each syntactic signature. This is a rephrasing of the
$\blank^{\A}$, $\blank^{\D}$ and $\blank^{\S}$ interpretations in
\cite{kaposi2019constructing}.

\begin{definition}[The $\Set$ model of $\ToS$] For each $j$, we have $\bA : \ToS_{j+1, j}$, which can be given by restricting the $\bM$ model
of Section \ref{sec:semantics} so that we only have the first $\Con$ components
in the interpretations for contexts, substitution, types, term, and we only have
actions on contexts in the interpretations of term and substitution
formers. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bA} && :\equiv \Set_{j+2} \\
  & \Sub_{\bA}\,\Gamma\,\Delta && :\equiv \Gamma \ra \Delta\\
  & \Ty_{\bA}\,\Gamma && :\equiv \Gamma \ra \Set_{j+2}\\
  & \Tm_{\bA}\,\Gamma\,A && :\equiv (\gamma : \Gamma) \ra A\,\gamma
\end{alignat*}
Now, for $\Gamma : \Con_{\syn}$, the type of $\Gamma$-algebras is given by
$\llb\Gamma\rrb_{\bA}$. E.g.\, $\llb\NatSig\rrb_{\bA}$ yields a left-nested
$\Sigma$-type of pointed sets with an endofunction. Also, $\llb\Gamma\rrb_{\bM}$
extends $\llb\Gamma\rrb_{\bA}$ to an flCwF of $\Gamma$-algebras.
\end{definition}

\begin{definition}[Logical predicate model of $\ToS$ over the $\Set$ model]
For each $j$ level we have $\bD$, which is a displayed $\ToS$ model over $\bA$
at level $j$. This model, analogously to $\bA$, is given by restricting $\bM$ to
the \emph{third} components everywhere, corresponding to types or actions on
types. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bD}\,\ulGamma && :\equiv \ulGamma \ra \Set_{j+2} \\
  & \Sub_{\bD}\,\Gamma\,\Delta\,\ulsigma && :\equiv (\ulgamma : \ulGamma) \ra \Gamma\,\gamma \ra \Delta(\llb\ulsigma\rrb_{\bA}\,\ulgamma)\\
  & \Ty_{\bD}\,\Gamma\,\ulA && :\equiv (\ulgamma : \ulGamma) \ra \Gamma\,\ulgamma \ra \llb\ulA\rrb_{\bA}\,\ulgamma \ra \Set_{j+2}\\
  & \Tm_{\bD}\,\Gamma\,A\,\ult && :\equiv (\ulgamma : \ulGamma)(\gamma : \Gamma\,\ulGamma)
    \ra A\,\ulgamma\,\gamma\,(\llb\ult\rrb_{\bA}\,\ulgamma)
\end{alignat*}
\end{definition}

For $\Gamma : \SynSig_j$, the type of displayed $\Gamma$-algebras over some
$\ulgamma : \llb\Gamma\rrb_{\bA}$ is given by $\llb\Gamma\rrb_{\bD}\,\ulgamma$. In
other words, $\llb\Gamma\rrb_{\bD}$ yields the notion of types in the flCwF of
$\Gamma$-algebras given by $\llb\Gamma\rrb_{\bM}$.

\begin{definition}[Displayed algebra section model of $\ToS$] Analogously to
$\bA$ and $\bD$, we define $\bS$ as a displayed $\ToS$ model over the total
model of $\bD$, which is given by restricting $\bM$ to the fourth components,
corresponding to interpretations of terms and actions on terms.

For $\ulgamma : \llb\Gamma\rrb_{\bA}$ and $\gamma : \llb\Gamma\rrb_{\bD}\,\ulgamma$,
the type of $\Gamma$-sections is computed as $\llb\Gamma\rrb_{\bS}\,\ulgamma\,\gamma$.
\end{definition}

@@

\begin{definition}[Term algebra construction]
For each syntactic signature $\ulOmega : \SynSig_j$, we define a displayed $\ToS$
model over $\syn_j$, named $\bT_{\ulOmega}$. The underlying sets are as follows:
\begin{alignat*}{3}
  & \Con_{\bT_{\ulOmega}}\,\ulGamma &&:\equiv \Sub\,\ulOmega\,\ulGamma \ra \llb\ulGamma\rrb_{\bA}\\
  & \Sub_{\bT_{\ulOmega}}\,\Gamma\,\Delta\,\ulsigma &&:\equiv (\ulnu : \Sub\,\ulOmega\,\ulGamma)
  \ra \Delta\,(\ulsigma\circ\nu) \simeq \llb\ulsigma\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \Ty_{\bT_{\ulOmega}}\,\Gamma\,\ulA &&:\equiv (\ulnu : \Sub\,\ulOmega\,\ulGamma) \ra
    \Tm\,\ulOmega\,(\ulA[\ulnu]) \ra \llb\ulA\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \Tm_{\bT_{\ulOmega}}\,\Gamma\,A\,\ult &&:\equiv (\ulnu : \Sub\,\ulOmega\,\ulGamma)\ra
    \llb\ulA\rrb_{\bA}\,\ulnu\,(\ult[\ulnu]) \simeq_{\id} \llb\ult\rrb_{\bA} (\Gamma\,\ulnu)
\end{alignat*}

Above, the $\simeq$ in the definition of $\Sub_{\bT_{\ulOmega}}$ is a context isomorphism
in $\llb\ulDelta\rrb_{\bM}$, which is the flCwF of $\ulDelta$-algebras. The
$\simeq_{\id}$ in $\Ty_{\bT_{\ulOmega}}$ is a vertical context isomorphism in the displayed
flCwF given by $\llb\ulA\rrb_{\bM}$.

So far, the underlying sets in $\bT_{\ulOmega}$ are similar to what was given in
\cite{kaposi2019constructing} in the construction of term algebras, but there is
an important difference: in ibid.\, strict equalities are used instead of
isomorphisms. In our case, isomorphisms are necessary once again because of
infinitary functions types and our identity type; we shall see this shortly. The
universe is interpreted as follows:
\begin{alignat*}{3}
  & \U_{\bT_{\ulOmega}} : (\ulnu : \Sub\,\ulOmega\,\ulGamma)(\ult : \Tm\,\ulOmega\,\U)\ra
              \Set_{j+1}\\
  & \U_{\bT_{\ulOmega}}\,\ulnu\,\ult :\equiv \Tm\,\ulOmega\,(\El\,\ult)\\
  & \El_{\bT_{\ulOmega}}\,a : (\ulnu : \Sub\,\ulOmega\,\ulGamma)
             (\ult : \Tm\,\ulOmega\,(\El\,(\ula[\ulnu])))\ra
              \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \El_{\bT_{\ulOmega}}\,a\,\ulnu\,\ult :\equiv (a\,\ulnu)\,\ult
\end{alignat*}

Hence, a syntactic $\ult : \Tm\,\ulOmega\,\U$ is interpreted as a set of terms
with type $\El\,\ult$. In the interpretation of $\El$, note that
\[
  a\,\ulnu : \U_{\bT_{\ulOmega}}\,\ulnu\,(\ula[\ulnu]) \simeq_{\id}
                 \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)
\]
hence
\[
a\,\ulnu : \Tm\,\ulOmega\,(\El\,(\ula[\ulnu])) \simeq_{\id}
           \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)
\]
The $\simeq_{\id}$ above is just an isomorphism of sets, since it lives in
$\llb\U\rrb_{\bM}$ which was given as the flCwF of sets in Section
\ref{sec:universe}. This above isomorphism is a good summary of the
construction: the interpretation of a $\ula : \Tm\,\ulOmega\,\U$ in the term
algebra is isomorphic to a set of terms.

Inductive functions are interpreted by transport along such isomorphism:
\[
  \Pii_{\bT_{\ulOmega}}\,a\,B\,\ulnu\,\ult :\equiv \lambda\,\alpha.\,
         B\,(\ulnu,\,(a\,\ulnu)^{-1}\,\alpha)\,(\ult\,\appitt\,((a\,\ulnu)^{-1}\,\alpha))
\]

For the infinitary function space, we need the following, where $\simeq_{\id}$ is
again set isomorphism.
\begin{alignat*}{3}
  & \Piinf_{\bT_{\ulOmega}}\,A\,b\,\ulnu : \Tm\,\ulOmega\,(\El\,(\Piinf\,A\,(\lambda\,\alpha.\,(b\,\alpha)[\ulnu])))\\
  & \hspace{4.2em}\simeq_{\id} ((\alpha : A) \ra \llb\alpha\rrb_{\bA}\,(\Gamma\,\ulnu))
\end{alignat*}
This can be given using the natural isomorphism consisting of $\appinf$ and
$\laminf$. However, the sides are not strictly equal. For the identity type, we
build the following isomorphism using equality reflection.
\begin{alignat*}{3}
  & \Id_{\bT_{\ulOmega}}\,a\,t\,u\,\ulnu : \Tm\,\ulOmega\,(\El\,(\Id\,(\ula[\ulnu])\,(\ult[\ulnu])\,(\ulu[\ulnu]))\\
  & \hspace{4.2em}\simeq_{\id} (\llb\ult\rrb_{\bA}\,(\Gamma\,\ulnu) =   \llb\ulu\rrb_{\bA}\,(\Gamma\,\ulnu))
\end{alignat*}

We omit the rest of the definition of $\bT_{\ulOmega}$. The interpretations of equations in
the CwF and fl-structure parts become fairly technical, and we also need to
utilize iso-cleaving to interpret type substitution and substitution
laws. However, the basic shape of the model remains similar to
\cite{kaposi2019constructing}.
\end{definition}

Now, we can build the term algebra for $\ulOmega$ by taking
$\llb\ulOmega\rrb_{\bT_{\ulOmega}}\,\id$, which has type $\llb\ulOmega\rrb_{\bA}$.

\emph{Remark.} If we start with a syntactic signature at level $j$, then the
underlying sets in the term algebra are all in $\Set_{j+1}$. Hence, the term
algebra for $\NatSig : \SynSig_0$ has an underlying set in $\Set_{j+1}$. This is
a bit inconvenient, since normally we would have natural number in $\Set_0$. Our
current term model construction cannot avoid this level bump, since $\syn_j$ is
necessarily a large type, and we do not have a way to construct a small set from
a large set of terms. Perhaps this would be possible with a \emph{resizing rule}
\cite{voevodsky2011resizing}. Also, if we only consider closed finitary QIITs,
with no possibility of referring to external types in signatures, then we can
modify the current term model construction so that we always build sets in
$\Set_0$. This would cover natural numbers and most dependent type theories.

\subsection{Term Algebras Support Induction}

%% We assume $\ulOmega : \SynSig_j$.
%% We abbreviate the term algebra
%% $\llb\ulOmega\rrb_{\bT_{\ulOmega}}\,\id$ as

%% $\llb\ulOmega\rrb_{\bT_{\ulOmega}}\,\id$ as b
%% $\llb\ulOmega\rrb_{\bT_{\ulOmega}}\,\id$ as

@@

%% write about limitations!
